Looking at Non-Linear Dimension Reductions as Models in the Data Space

### 2025-06-11 paper.qmd
Finalize American English spellings in the content

- Remove all the extraneous commented text (Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible (changed in Rendering the model)
- Check spelling is consistently American
- Check references (Done)
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place.
- Move the notation table to appendix (Done)
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.

@@

The digit "1" of the MNIST dataset [@lecun1998] consists of $7877$ grayscale images of handwritten "1"s. Each image is $28 \times 28$ pixels which corresponds to 784 variables. The first 10 principal components, explaining 83% of the total variation, are used. This data is essentially lies on a nonlinear manifold in the high dimensions, defined by the shapes that "1"s make in when sketched, menaing that some pixels are almost always dark and some almost always light. We expect that the best layout captures this type of structure and does not exhibit distinct clusters.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)~parameter choice(s). It depends on conceptualising the \gD{} layout as a model, creating a wireframe representation of it that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.

Two examples illustrating usage are provided: the PBMC3k data where the NLDR is summarising clustering in \pD{} and hand-written digits illustrating how NLDR represents an intrinsically lower dimensional nonlinear manifold. We demonstrated that the published layout of the PBMC3k is inaccurate, because it grossly exaggerates separation between clusters, and even suggests separation when there is none. This is common when layouts are chosen subjectively -- often a preference for the "prettiest". Our approach provides a way to objectively choose the layout and hopefully avoids the use of misleading layouts in the future. In the hand-written digits we illustrate how our model fit statistics show that a flat disc layout is superior to the curved shaped layouts, and how to identify oddly written "1"s using the residuals of the fitted model.

>>>>

- Remove all the extraneous commented text (Done)
- Check how \kD{} and \gD{} are used, and make sure it's sensible (changed in Rendering the model: Done)
- Check spelling is consistently American (Done)
- Check references (Done)
- Add videos - make a list of the videos in the supplementary section, with the number of the figure that it corresponds too. This avoids having captions that are too long and awkward with raw URLs. And it keeps all of this supplementary material in one place.
- Move the notation table to appendix (Done)
- Code clean up: recommend that all of the calculations done with quollr are done with a separate script(s) and put in a scripts directory. Save the resulting data need to make plots and tables. Only code for the plots and tables is directly in the qmd in the final version.

@@

The digit "1" of the MNIST dataset [@lecun1998] consists of $7877$ grayscale images of handwritten "1"s. Each image is $28 \times 28$ pixels which corresponds to $784$ variables. The first $10$ principal components, explaining $83\%$ of the total variation, are used. This data is essentially lies on a nonlinear manifold in the high dimensions, defined by the shapes that "1"s make in when sketched, meaning that some pixels are almost always dark and some almost always light. We expect that the best layout captures this type of structure and does not exhibit distinct clusters.

@@

We have developed an approach to help assess and compare NLDR layouts, generated by different methods and (hyper-)~parameter choice(s). It depends on conceptualizing the \gD{} layout as a model, creating a wireframe representation of it that can be lifted into \pD{}. The fit is assessed by viewing the model in the data space, computing residuals and RMSE. Different layouts can be compared using the RMSE, and provides quantitative and objective methods for deciding on the most suitable NLDR layout to represent the \pD{} data. It also provides a way to predict the values of new \pD{} observations in the \gD{}, which could be useful for implementing uncertainty checks such as using training and testing samples.

Two examples illustrating usage are provided: the PBMC3k data where the NLDR is summarizing clustering in \pD{} and hand-written digits illustrating how NLDR represents an intrinsically lower dimensional nonlinear manifold. We demonstrated that the published layout of the PBMC3k is inaccurate, because it grossly exaggerates separation between clusters, and even suggests separation when there is none. This is common when layouts are chosen subjectively -- often a preference for the "prettiest". Our approach provides a way to objectively choose the layout and hopefully avoids the use of misleading layouts in the future. In the hand-written digits we illustrate how our model fit statistics show that a flat disc layout is superior to the curved shaped layouts, and how to identify oddly written "1"s using the residuals of the fitted model.
